\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-12 doc/publications/plos16/plos16.tex
changes to  plos16.tex

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\textbf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

@@

\marginpar{make consistent capitalization for ranger, randomForest}

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest\ outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
trees and use a high \mtry value which leads to greater accurace (see section \ref{section:synthetic}). 
\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
We do not see that}

@@

\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

In the first section we explore some of the properties of random forests as these will also be apparent in the
\cursedforest\ implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
of \cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
the 1000 Genomes Project. We report the error in predicting ethnicity.  Finally, given the role
different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
prediction accuracy of the model.

@@

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can still simply schedule extra tasks to handle the additional items in the RDD.

@@

On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
work down to the level of the individual features.  For example, for each node of a tree, \cursedforest will distribute
tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
entire decision tree.

The current implementation of \cursedforest uses an ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
\[
I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 − f_q ), 
\]
which is a minimum when all observatkions at the node are in the same class.

@@

\subsubsection{Synthetic data} 
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.

The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ then 
\begin{align*}
z=& \Sigma_{i=1}^{i=5} {w_i x_i} \\
\intertext{and}
\sigma_\epsilon^2 =& Var(z)(1- 0.125)/0.125 \\
\intertext{so}
y =&  z+\epsilon \textrm{ where } \epsilon \sim  N(0, \sigma_\epsilon^2).
\end{align*}
The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \randomforest by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate two features of \cursedforest\, a variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
samples, as discussed in the methods section.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
classification error and the rank-biased-overlap
\marginpar{what is wrong with the formatting of this citation?}
\cite[]{Webber.et.al.2010} measure.

@@

% From the synthetic data we can conclude that \cursedforest\ is able to perform accurate feature selection as well as
% accurate classification for large numbers of samples with large feature vectors.

\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
Fig.~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
the default. Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
values of \mtry and \ntree considered  whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is less
sensitive once we over a certain threshold.

From the synthetic data we can conclude that \cursedforest\ is able to perform accurate feature selection as well as
accurate classification for large numbers of samples with large feature vectors.

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{ out of bag classification error and Rank-biased overlap estimates}} 
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
      \caption{OOB classification error} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo.png}
      \caption{The  rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
      the better result, that is, a lower OOB or a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated cancer variants}
We consider the question of feature discovery on the Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
VCF files retrieved from the TCGA data portal.  For the label, we use the cell type, tumor or normal.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ outperforms existing methods for classification problems}
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
1000 Genomes Project dataset.

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\centering
\caption{ {\bf Performance comparison with \cursedforest\ on different subsets of the 1000 Genomes Project data.}}
\begin{tabular}{| l | r | l |}
\hline
{\bf Genome \%} & {\bf Error} & {\bf Runtime} \\ \thickhline
phase1\_chr22: 1,092 x 490,036 & 0.04 & 1 min 5 seconds \\ \hline
phase3\_chr22: 2,504 x 1,103,548 & 0.06 & 1 min 37 seconds \\ \hline
phase3\_chr1: 2,504 x 6,450,364 & 0.04 & 4 min 10 seconds \\ \hline
phase3\_chr1-3: 2,504 x 19,328,051 & 0.02 & 10 min 49 seconds \\ \hline
phase3\_chr1-22: 2,504 x 81,047,467 & 0.01 & 36 min 54 seconds \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of runtime and error for models built using \cursedforest\ on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees and was built using 130 worker nodes---each with 8GB RAM.
\end{flushleft}
\label{cursedforesttable}
\end{minipage}
\end{table}

@@

\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\
                                                & \randomforest (R)         & 0.05       & 24 min           \\
                                                & \ranger (R)      & 0.06       & 1 min 10 seconds       \\
                                                & \ranger (C++)     & 0.06       & 7 min            \\
                                                & H2O           & ??       & ??         \\

@@

\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest\ & 0.06 & 4 min 2 seconds \\
                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)    & 0.06       & 4 min 30 seconds     \\
                                                    & \ranger (C++)     &     ??   & ??        \\
                                                    & H2O           & 0.05       & 16 hours         \\

@@

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.04  & 19 min 19 seconds         \\
                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)      &  0.05       & 28 mins         \\
                                                    & \ranger (C++)     & ??       & ??            \\
                                                    & H2O           & NA       & NA         \\
\hline

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.02  & 56 min 43 seconds             \\
                                                    & Spark ML &    NA         &   NA               \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)      & 0.03      &    3 hour 58 min     \\
                                                    & \ranger (C++)     & ??       & ??            \\
                                                    & H2O           & NA       & NA         \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.01  & 3hour 16min \\
                                                    & Spark ML & NA & NA  \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)       &        NA     &        NA    \\
                                                    & \ranger (C++)       &        NA     &        NA    \\
                                                    & H2O           & NA       & NA         \\
\hline

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 4: Sensitivity 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scalability analysis}
In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data
and computational resources.

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\ranger, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
of random forests by using a range of \mtry values.

We investigate the different synthetic datasets generated for section 5.1. and measured the time taken to build a random
forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases were executed with the
same random seed, to improve the consistency of measurements.

First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. Regardless of the
number of cores used, \cursedforest displays approximately linear dependency between the execution time and \mtry
(Fig. \ref{figure:synthetictiming.a}).

@@

Next we investigate \cursedforest scalability with regards to the size of data, by varying the number of variables and
sample for a fixed \mtry fraction of 0.25 and execution of 128 CPU cores. The raw results are presented in the table
[REF] 
\marginpar{have we lost this table?}
and visualized in figure \ref{figure:synth2} below (please note log scale on the axes and the values on y axes are
expresses as trees per hour).

@@

To conclude \cursedforest\ is capable of processing datasets with 50M variables and 10K samples, which is the order of
magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per
hour.

@@

%nvars 150000  500000  2500000 10000000  50000000
%nsamples          
%1000  21.6  26.7  59.8  133.5 668.8
%5000  45.7  71.9  186.2 729.3 2887.4
%10000 70.2  120.2 397.3 2237.8  5747.8

\begin{table}[!ht]
\begin{adjustwidth}{-1.00in}{0in}
\centering
\caption{\textbf{
 Time in seconds to build a random forest of 100
  trees on 128 CPU cores with varying numbers of variables and samples.}}
\label{synthetictimingtable}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
\cline{3-7}
\multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
\hline                                                            
\multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
                                        & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
                                        & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
\hline
\end{tabular}
\end{adjustwidth}
\end{table}

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\textbf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

@@

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
trees and use a high \mtry value which leads to greater accuracy (see the example in section \ref{section:synthetic}). 
\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
We do not see that. }

@@

\cursedforest is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. This offers a
comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
utilized to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

In the first section we explore some of the properties of random forests as these will also be apparent in the
\cursedforest implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
of \cursedforest in respect to the dimensionality of the data, building a random forest model on whole-genome data from
the 1000 Genomes Project. We report the error in predicting ethnicity.  Finally, given the role
different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
prediction accuracy of the model.

@@

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.

@@

On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
work down to the level of the individual features.  For each node of a tree, \cursedforest will distribute
tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
entire decision tree.

The current implementation of \cursedforest uses an ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 − f_q ), 
$$
which is a minimum when all observatkions at the node are in the same class.

@@

\subsubsection{Synthetic data} 
\label{section:synthetic_data}
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.

The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ and we set 
$$
z =  \sum_{i=1}^{i=5} {w_i x_i}.
$$
We let $\sigma_\epsilon^2 = Var(z)(1- 0.125)/0.125 $ and then 
$y =   z + \epsilon$ where $\epsilon \sim  N(0, \sigma_\epsilon^2).$
The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the boostrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.} In the TCGA example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate two features of \cursedforest, a variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
samples, as discussed in the methods section.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
classification error and the rank-biased-overlap (RBO)
\marginpar{what is wrong with the formatting of this citation?}
\cite[]{Webber.et.al.2010} measure.

@@

% From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
% accurate classification for large numbers of samples with large feature vectors.

\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
Fig.~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
the default. Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
values of \mtry and \ntree considered  whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
sensitive once we over a certain threshold.

From the synthetic data we can conclude that \cursedforest is able to perform accurate feature selection as well as
accurate classification for large numbers of samples with large feature vectors.

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
      \caption{OOB classification error.} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo.png}
      \caption{The rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
      the better result, that is, a lower OOB or a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest selects previously demonstrated cancer variants}
We consider the question of feature discovery on the Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
VCF files retrieved from the TCGA data portal.  For the label, we use the cell type, tumor or normal.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest outperforms existing methods for classification problems}
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
1000 Genomes Project dataset.

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\centering
\caption{ {\bf Performance comparison with \cursedforest on different subsets of the 1000 Genomes Project data.}}
\begin{tabular}{| l | r | l |}
\hline
{\bf Genome \%} & {\bf Error} & {\bf Runtime} \\ \thickhline
phase1\_chr22: 1,092 x 490,036 & 0.04 & 1 min 5 seconds \\ \hline
phase3\_chr22: 2,504 x 1,103,548 & 0.06 & 1 min 37 seconds \\ \hline
phase3\_chr1: 2,504 x 6,450,364 & 0.04 & 4 min 10 seconds \\ \hline
phase3\_chr1-3: 2,504 x 19,328,051 & 0.02 & 10 min 49 seconds \\ \hline
phase3\_chr1-22: 2,504 x 81,047,467 & 0.01 & 36 min 54 seconds \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of runtime and error for models built using \cursedforest on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees and was built using 130 worker nodes---each with 8GB RAM.
\end{flushleft}
\label{cursedforesttable}
\end{minipage}
\end{table}

@@

\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest & 0.04  & 1 min 50 seconds                        \\
                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\
                                                & \randomforest (R)         & 0.05       & 24 min           \\
                                                & \ranger (R)      & 0.06       & 1 min 10 seconds       \\
                                                & \ranger (C++)     & 0.06       & 7 min            \\
                                                & H2O           & ??       & ??         \\

@@

\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest & 0.06 & 4 min 2 seconds \\
                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)    & 0.06       & 4 min 30 seconds     \\
                                                    & \ranger (C++)     &     ??   & ??        \\
                                                    & H2O           & 0.05       & 16 hours         \\

@@

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest & 0.04  & 19 min 19 seconds         \\
                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)      &  0.05       & 28 mins         \\
                                                    & \ranger (C++)     & ??       & ??            \\
                                                    & H2O           & NA       & NA         \\
\hline

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest & 0.02  & 56 min 43 seconds             \\
                                                    & Spark ML &    NA         &   NA               \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)      & 0.03      &    3 hour 58 min     \\
                                                    & \ranger (C++)     & ??       & ??            \\
                                                    & H2O           & NA       & NA         \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest & 0.01  & 3hour 16min \\
                                                    & Spark ML & NA & NA  \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)       &        NA     &        NA    \\
                                                    & \ranger (C++)       &        NA     &        NA    \\
                                                    & H2O           & NA       & NA         \\
\hline

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 4: Sensitivity 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scalability analysis}
In this section we explore the performance of \cursedforest by testing its ability to scale to different sizes of data
and computational resources.

In order to asses these characteristics we ran \cursedforest classification on synthetic datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\ranger, allocating varying number of CPU cores to the \cursedforest and also varying the computational complexity
of the random forests by using a range of \mtry values.

We investigate the different synthetic datasets generated for section \ref{ection:synthetic_data} and measured the time
taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases
were executed with the same random seed, to improve the consistency of measurements.

First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. 
The results are  visualized in figure \ref{figure:synthetictiming} below. 
%The raw results are
%presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. 
Regardless of the number of cores used, \cursedforest displays approximately linear dependency between the execution
time and \mtry (Fig. \ref{figure:synthetictiming.a}).

@@

\marginpar{tables 2 has been commented out -- there is no data in it. Piotr could add the data or we could remove both
  tables just leave the figures. I have done this for the time being} 
Next we investigate \cursedforest scalability with
regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
execution of 128 CPU cores. The results are visualized in figure \ref{figure:synth2} below (please note log scale on the
axes and the values on y axes are expresses as trees per hour).

@@

To conclude \cursedforest is capable of processing datasets with 50M variables and 10K samples, which is the order of
magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per
hour.

@@

% commented this table out -- see above
% \begin{table}[!ht]
%   \begin{adjustwidth}{-1.00in}{0in}
%     \centering
%     \caption{\textbf{
%  Time in seconds to build a random forest of 100
%   trees on 128 CPU cores with varying numbers of variables and samples.}}
% \label{synthetictimingtable}
% \begin{tabular}{|l|l|l|l|l|l|l|}
% \hline
% \multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
% \cline{3-7}
% \multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
% \hline                                                            
% \multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
%                                         & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
%                                         & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
% \hline
% \end{tabular}
% \end{adjustwidth}
% \end{table}
